---
output:
  rmarkdown::github_document:
    html_preview: no
---

<!--- README.md is generated from README.Rmd. Please edit that file -->

# robust.prioritizr

## Spatial conservation planning under uncertainty

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/Lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/actions/workflow/status/frankiecho/roubust.prioritizr/R-CMD-check-ubuntu.yaml?branch=main&label=Ubuntu)](https://github.com/frankiecho/roubust.prioritizr/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/actions/workflow/status/frankiecho/roubust.prioritizr/R-CMD-check-windows.yaml?branch=main&label=Windows)](https://github.com/frankiecho/roubust.prioritizr/actions)
[![R-CMD-check-macOS](https://img.shields.io/github/actions/workflow/status/frankiecho/roubust.prioritizr/R-CMD-check-macos.yaml?branch=main&label=macOS)](https://github.com/frankiecho/roubust.prioritizr/actions)
[![Documentation](https://img.shields.io/github/actions/workflow/status/frankiecho/roubust.prioritizr/documentation.yaml?branch=main&label=Documentation)](https://github.com/frankiecho/roubust.prioritizr/actions)
[![Coverage-Status](https://img.shields.io/codecov/c/github/frankiecho/roubust.prioritizr?label=Coverage)](https://app.codecov.io/gh/frankiecho/roubust.prioritizr/branch/main)
[![CRAN-Status-Badge](https://www.r-pkg.org/badges/version/robust.prioritizr)](https://CRAN.R-project.org/package=robust.prioritizr)

<!-- badges: end -->

```{r, include = FALSE}
devtools::load_all()
plot <- function(...) terra::plot(..., cex = 1.5)
```

```{r, include = FALSE}
# set parameters for figure sizes
h <- 4.5
w <- 6.5
ow <- "500"
knitr::opts_chunk$set(
  fig.height = h,
  fig.width = w,
  out.width = ow,
  fig.path = "man/figures/README-", fig.align = "center"
)
```

```{r, include = FALSE}
# set up print method
print <- function(x, ...) {
  if (inherits(x, "ConservationProblem")) {
    prioritizr::knit_print.ConservationProblem(x)
  } else if (inherits(x, "OptimizationProblem")) {
    prioritizr::knit_print.OptimizationProblem(x)
  } else {
    base::print(x)
  }
}
```

TODO

## Installation

#### Developmental version

The latest development version can be installed to gain access to new functionality that is not yet present in the latest official version. Please note that the developmental version is more likely to contain coding errors than the official version. To install the developmental version, you can install it directly from the [GitHub online code repository](https://github.com/frankiecho/roubust.prioritizr) with the following _R_ code.

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("frankiecho/roubust.prioritizr")
```

## Citation

TODO.

## Usage

TODO.

## Learning resources

TODO.

## Getting help

If you have any questions about the _robust.prioritizr R_ package or suggestions for improving it, please [post an issue on the code repository](https://github.com/frankiecho/roubust.prioritizr/issues/new).
